Load tidyverse libraries.

```{r,message=FALSE,warnings=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
```

Read in data, which is just the Excel sheet saved as a CSV.

```{r}
data <- read.csv("Data624_project1_data.csv",header=TRUE)
```

Remove last rows where all fields are blank.

```{r}
data <- data[1:9732,]
```

Convert to long format.

```{r}
data_gathered <- gather(data,
	key="Variable",
	value="Value",
	-SeriesInd,-group)
data_gathered <- data.frame(data_gathered,
	Group.plus.var = paste0(data_gathered$group,"\n",data_gathered$Variable),
	stringsAsFactors=FALSE)
data_gathered$Group.plus.var <- factor(data_gathered$Group.plus.var,
	levels=paste0(rep(paste0("S0",1:6),each=5),"\n",rep(c("Var01","Var02","Var03","Var05","Var07"),times=6)))
```

Make line plots.

```{r,fig.width=12,fig.height=8}
ggplot(data_gathered,
	aes(SeriesInd,Value)) +
	geom_line() +
	facet_wrap(~Group.plus.var,scales="free_y",nrow=6,ncol=5) +
	theme(axis.text.x=element_text(angle=90, hjust=1))
```

Plot minus S06 consistent outliers (SeriesInd 41128, variables 1,3, 5 and 7) and minus S02 consistent outliers (SeriesInd 42951, variables 1, 3, 5, and 7).

```{r}
extreme_outliers <- which(data_gathered$SeriesInd == 41128 & data_gathered$group == "S06" & data_gathered$Variable != "Var02")
extreme_outliers <- c(extreme_outliers,
	which(data_gathered$SeriesInd == 42951 & data_gathered$group == "S02" & data_gathered$Variable != "Var02"))
data_gathered_minus_outliers <- data_gathered[setdiff(1:nrow(data_gathered),extreme_outliers),]
```

```{r,fig.width=12,fig.height=8}
ggplot(data_gathered_minus_outliers,
	aes(SeriesInd,Value)) +
	geom_line() +
	facet_wrap(~Group.plus.var,scales="free_y",nrow=6,ncol=5) +
	theme(axis.text.x=element_text(angle=90, hjust=1)) +
	ggtitle("Minus S06 SeriesInd 41128 and S02 SeriesInd 42951 in vars 1/3/5/7")
```

We now see more clearly that variables 1, 3, 5, and 7 tend to have a similar time pattern within each series.

Let's try plotting these variables on the same plot for each series.

```{r,fig.width=12,fig.height=8}
ggplot(data_gathered_minus_outliers[data_gathered_minus_outliers$Variable != "Var02",],
	aes(SeriesInd,Value,colour=Variable)) +
	geom_line() +
	facet_wrap(~group,scales="free_y",nrow=3,ncol=2) +
	theme(axis.text.x=element_text(angle=90, hjust=1)) +
	ggtitle("Minus S06 SeriesInd 41128 and S02 SeriesInd 42951 in vars 1/3/5/7")
```

Now, plot Var02 alone.

```{r,fig.width=12,fig.height=8}
ggplot(data_gathered[data_gathered$Variable == "Var02",],
	aes(SeriesInd,Value)) +
	geom_line() +
	facet_wrap(~group,scales="free_y",nrow=3,ncol=2) +
	theme(axis.text.x=element_text(angle=90, hjust=1)) +
	ggtitle("Var02")
```

Next, let's see if there are any missing values in the data.

```{r}
length(which(is.na(data_gathered) == TRUE))

data_gathered <- data_gathered[order(data_gathered$SeriesInd,data_gathered$group),]

table(data_gathered[which(is.na(data_gathered$Value) == TRUE),"SeriesInd"])

data_gathered[which(is.na(data_gathered$Value) == TRUE & (data_gathered$SeriesInd == 40697 | data_gathered$SeriesInd == 41821)),]
data_gathered[which(is.na(data_gathered$Value) == TRUE & (data_gathered$SeriesInd == 42897 | data_gathered$SeriesInd == 42898)),]
data_gathered[which(is.na(data_gathered$Value) == TRUE & (data_gathered$SeriesInd == 42997 | data_gathered$SeriesInd == 43000)),]
```

We find the following missing values.

* SeriesInd 40697, all five variables of S06. Simply remove or use the nearby timepoints within the same series and variable to fill in.
* SeriesInd 41821, all five variables of S05. Simply remove or use the nearby timepoints within the same series and variable to fill in.
* SeriesInd 42897 and 42898, all six series and all four correlated variables (1/3/5/7). Simply remove or use the nearby timepoints within the same series and variable to fill in.
* SeriesInd 42997 and 43000, all six series and 3/4 correlated variables (3/5/7). Can either use nearby timepoints or the correlated variable without missing values (Var01).

