---
title: "H+A Ch. 7: Exponential Smoothing"
author: "Heather Geiger"
output:
  html_document:
    code_folding: hide
---	

# Libraries

Load all libraries needed to complete this analysis.

```{r, message=FALSE,warning=FALSE}
library(fpp2)
library(ggplot2)
library(tidyr)
library(dplyr)
```

# H+A Question 7.5

## Question

Date set **books** contains the daily sales of paperback and hardcover books at the same store. The task is to forecast the next four days’ sales for paperback and hardcover books.

1. Plot the series and discuss the main features of the data.
2. Use the **ses()** function to forecast each series, and plot the forecasts.
3. Compute the RMSE values for the training data in each case.

## Answer

### Part 1 - Exploratory analysis

Start with a simple line plot.

```{r}
books_dat <- data.frame(Day = rep(1:30,times=2),
	Type = rep(c("Paperback","Hardcover"),each=30),
	Sales = c(books[,1],books[,2]),
	stringsAsFactors=FALSE)

ggplot(books_dat,
aes(Day,Sales)) +
geom_line() +
facet_wrap(~Type)
```

It feels like there could potentially be a day of the week pattern here, but it's a bit hard to tell from this plot.

Let's plot each set of 7 days in its own panel.

```{r}
books_dat <- data.frame(books_dat,
	Day.of.week = rep(c(rep(1:7,times=4),1,2),times=2),
	Week = paste0("Week",rep(c(rep(1:4,each=7),5,5),times=2)),
	stringsAsFactors=FALSE)

ggplot(books_dat,
aes(Day.of.week,Sales)) +
geom_line() +
geom_point() +
facet_grid(Type ~ Week) +
xlab("Day of the week")
```

We don't know here which actual day of the week from Sunday to Saturday each day of the week is, but we could infer if there were a weekday vs. weekend pattern if we saw two consecutive days consistently high/low. 

Or if there were one particular day of the week with higher sales for whatever reason, we would be able to see it from this plot.

Here, it really does not seem like there is any pattern within what day of the week it is.

As for trend, it seems like there could possibly be an increasing trend, though it's admittedly a bit hard to tell with such strong daily fluctuations. For our forecasting, we'll move forward based on the assumption that there is no trend and just run simple exponential smoothing.

### Part 2 - Forecasting with ses

Use the ses function to forecast.

```{r}
paperback_forecast <- ses(books[,1],h=4)
hardcover_forecast <- ses(books[,2],h=4)

autoplot(paperback_forecast) + 
xlab("Day") + ylab("Sales") + 
ggtitle("Paperback sales from past 30 days\n+ forecast through next 4 days (with confidence intervals)")

autoplot(hardcover_forecast) +
xlab("Day") + ylab("Sales") + 
ggtitle("Hardcover sales from past 30 days\n+ forecast through next 4 days (with confidence intervals)")
```

Print the numbers for each.

```{r, echo=FALSE}
print("Paperback forecast:")
```

```{r}
paperback_forecast
```

```{r, echo=FALSE}
print("Hardcover forecast:")
```

```{r}
hardcover_forecast
```

### Part 3 - Compute RMSE values for the training data in each case.

The MSE (mean squared error) is already included in "mse" within the "model" from the forecast object.

Just take the square root of this to get the RSME (root mean squared error).

```{r, echo=FALSE}
print("RMSE paperback forecast:")
```

```{r}
sqrt(paperback_forecast$model$mse)
```

```{r, echo=FALSE}
print("RMSE hardcover forecast:")
```

```{r}
sqrt(hardcover_forecast$model$mse)
```

The RSME can be thought of as something similar to the standard deviation for the residuals. Basically, it's a way to measure the spread of how much the actual values deviate from the predictions of the model within the training data.

Here, both paperback and hardcover forecasts have relatively similar RSME values of ~32-34. If the residuals are normally distributed, this means we would expect most predictions to fall within +/- 32-34 books vs. the actual value. While the vast majority would fall within +/- 64-68 books.

# H + A Question 7.6

## Question

This question uses the same data as question 7.5.

1. Now apply Holt’s linear method to the **paperback** and **hardcover** series and compute four-day forecasts in each case.
2. Compare the RMSE measures of Holt’s method for the two series to those of simple exponential smoothing in the previous question. (Remember that Holt’s method is using one more parameter than SES.) Discuss the merits of the two forecasting methods for these data sets.
3. Compare the forecasts for the two series using both methods. Which do you think is best?
4. Calculate a 95% prediction interval for the first forecast for each series, using the RMSE values and assuming normal errors. Compare your intervals with those produced using **ses** and **holt**.

## Answers

### Part 1 - Apply Holt's linear method to both series and compute four-day forecasts in each case.

```{r}
paperback_holt <- holt(books[,1],h=4)
hardcover_holt <- holt(books[,2],h=4)

autoplot(paperback_holt) +
xlab("Day") + ylab("Sales") + 
ggtitle("Paperback sales from past 30 days\n+ Holt's linear trend forecast through next 4 days")

autoplot(hardcover_holt) +
xlab("Day") + ylab("Sales") + 
ggtitle("Hardcover sales from past 30 days\n+ Holt's linear trend forecast through next 4 days")
```

```{r, echo=FALSE}
print("Paperback forecast using Holt's linear trend:")
```

```{r}
paperback_holt
```

```{r, echo=FALSE}
print("Hardcover forecast using Holt's linear trend:")
```
```{r}
hardcover_holt
```

### Part 2 - Compare RSME measures vs. simple exponential smoothing.

First, calculate RSME for Holt's forecasts.

```{r, echo=FALSE}
print("RSME for Holt's on paperback")
```

```{r}
sqrt(paperback_holt$model$mse)
```

```{r, echo=FALSE}
print("RMSE for Holt's on hardcover")
```

```{r}
sqrt(hardcover_holt$model$mse)
```

The RMSE error values using Holt's linear trend is lower for both series than using simple exponential smoothing.

Assuming both methods have a mean of the residuals around 0, this would mean that we should prefer Holt's linear trend, since its predictions show less deviation from being perfectly accurate.

However, this is a big assumption! Let's check if it's true.

```{r, echo=FALSE}
print("Means of residuals for SES, then Holt's, for paperback series:")
```

```{r}
mean(paperback_forecast$residuals)
mean(paperback_holt$residuals)
```

```{r, echo=FALSE}
print("Means of residuals for SES, then Holt's, for hardcover series:")
```

```{r}
mean(hardcover_forecast$residuals)
mean(hardcover_holt$residuals)
```

Looks like the mean of the residuals using Holt's is actually closer to 0 than for SES, so that's another check in favor of Holt's.

Finally, let's visualize the distribution of residuals as well as plot time vs. residuals.

```{r}
par(mfrow=c(2,2))

hist_residuals <- function(residuals,title){
	hist(residuals,
		xlab="Residuals",
		ylab="Number of predictions",
		labels=TRUE,
		breaks=seq(from=-80,to=80,by=20),
		main=title)
}

hist_residuals(paperback_forecast$residuals,
	"Paperback, SES")
hist_residuals(paperback_holt$residuals,
	"Paperback, Holt's linear")
hist_residuals(hardcover_forecast$residuals,
	"Hardcover, SES")
hist_residuals(hardcover_holt$residuals,
	"Hardcover, Holt's linear")
```

```{r}
residuals_vs_time <- function(residuals,title){
	plot(1:30,
		residuals,
		xlab="Day",
		ylab="Residuals",
		type="p",
		main=title,
		ylim=c(-70,80))

abline(h=0,lty=2)
}

par(mfrow=c(2,2))

residuals_vs_time(paperback_forecast$residuals,
	"Paperback, SES")
residuals_vs_time(paperback_holt$residuals,
	"Paperback, Holt's linear")
residuals_vs_time(hardcover_forecast$residuals,
	"Hardcover, SES")
residuals_vs_time(hardcover_holt$residuals,
	"Hardcover, Holt's linear")
```

The distributions of residuals look pretty similar. It's hard to tell with only 30 data points, but they all seem relatively close to normal.

There does not seem to be any major bias across time for predictions to be in any one direction.

Therefore with these other factors being equal and RSME being lower using Holt's linear trend, it seems that Holt's linear trend has the clear advantage if we consider only performance within the training data.

### Part 3 - Compare the forecasts for the two series using both methods. Which do you think is best?

We decided in part 2 that Holt's linear trend would definitely be considered best if we looked only at performance within the training data.

Now, we are asked to also compare them using the forecasts into the future.

We saw above that SES gives the same forecast value for all four days into the future, with only the prediction intervals changing over time. Whereas Holt's linear trend gives a different (in this case increasing) value for each point further out in time.

Paperback SES gave a prediction of 207.1097 books for all four days.

Meanwhile Holt's method gave a prediction pretty close to the prediction from SES (209.4668 books) on the first day. But by the last day its prediction was decently higher (213.2197 books).

Similar idea for the hardcover series (239.5601 from SES, 250.1739 to 260.0817 from Holt's linear method).

Let's look at how many days had sales higher than these different predictions within each series.

```{r, echo=FALSE}
print("Days with sales higher than SES prediction, then first day Holt's prediction, then last day Holt's prediction, paperback series:")
```

```{r}
length(which(books[,1] > paperback_forecast$mean[1]))
length(which(books[,1] > paperback_holt$mean[1]))
length(which(books[,1] > paperback_holt$mean[4]))
```

```{r, echo=FALSE}
print("Same for hardcover series:")
```

```{r}
length(which(books[,2] > hardcover_forecast$mean[1]))
length(which(books[,2] > hardcover_holt$mean[1]))
length(which(books[,2] > hardcover_holt$mean[4]))
```

We find that both SES and Holt's method predict future sales that are higher than most sales of the past 30 days, for both series. So while the predictions using Holt's method are even higher than for SES, it's not like they are that much more out of the realm than the predictions from SES.

With this information combined with the better performance within the training data, I would continue to say that Holt's linear trend is the better choice over SES for forecasting from this data.

### Part 4 - Calculate prediction intervals and compare to intervals from SES and Holt functions.

First, let's review the intervals from SES and Holt functions.

Prediction interval for first forecast from SES, paperback series: 138.8670-275.3523 (range ~136)

Holt's linear trend, paperback series: 143.9130-275.0205 (range ~131)

SES, hardcover series: 174.7799-304.3403 (range ~130)

Holt's linear trend, hardcover series: 192.9222-307.4256 (range ~115)

Now, compare to what we would get from calculating it ourselves using RMSE and assumption of a normal distribution.

```{r}
zscore_for_95percent_interval <- qnorm(.975)
```

```{r, echo=FALSE}
print("Intervals for SES then Holt's, paperback series:")
```

```{r}
plus_minus_interval <- sqrt(paperback_forecast$model$mse)*zscore_for_95percent_interval

c(paperback_forecast$mean[1] - plus_minus_interval,paperback_forecast$mean[1] + plus_minus_interval)

plus_minus_interval <- sqrt(paperback_holt$model$mse)*zscore_for_95percent_interval

c(paperback_holt$mean[1] - plus_minus_interval,paperback_holt$mean[1] + plus_minus_interval)
```

```{r,echo=FALSE}
print("Intervals for SES then Holt's, hardcover series:")
```

```{r}
plus_minus_interval <- sqrt(hardcover_forecast$model$mse)*zscore_for_95percent_interval

c(hardcover_forecast$mean[1] - plus_minus_interval,hardcover_forecast$mean[1] + plus_minus_interval)

plus_minus_interval <- sqrt(hardcover_holt$model$mse)*zscore_for_95percent_interval

c(hardcover_holt$mean[1] - plus_minus_interval,hardcover_holt$mean[1] + plus_minus_interval)
```

To summarize again:

SES, paperback: 138.8670-275.3523 from function vs. 141.1810-273.0383 calculated

Holt's, paperback: 143.9130-275.0205 vs. 148.4395-270.4940

SES, hardcover: 174.7799-304.3403 vs. 176.9765-302.1437

Holt's, hardcover: 192.9222-307.4256 vs. 196.8754-303.4723

All of our calculated intervals are very slightly narrower than those given by the forecasting functions.

Overall though, they are quite similar.
