---
title: "Data 624, Project 1, Group 1: Exploration and forecasting of select series"
author: "Heather Geiger"
output: 
 html_document:
  code_folding: hide
  smart: false
---

# Introduction

From the assignment instructions:

"Your data is a de-identified Excel spreadsheet.  Your assignment is to perform the appropriate analysis to forecast several series for 140 periods.  You will have 1622 periods for your analysis."

I saved the Excel spreadsheet in a CSV file, and will be using this for all analysis.

We also divided the work so that each group member would analyze three series. My assigned series are S02, S03, and S04. Full forecast list:

* S02 – Forecast  Var02, Var03
* S03 – Forecast  Var05, Var07
* S04 – Forecast  Var01, Var02

# Libraries

Load in relevant libraries, including tidyverse and the forecasting library fpp2.

Also lubridate for easier handling of dates, and imputeTS to fill in NAs in a time series.

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(fpp2)
library(lubridate)
library(imputeTS)
```

# Data exploration

## Read in and format data.

Read in the data from the CSV file.

```{r}
data <- read.csv("data624_project1_excel_sheet_as_csv.csv",
	header=TRUE,stringsAsFactors=FALSE)
```

Convert SeriesInd from Excel date number format to regular R dates.

```{r}
data$SeriesInd <- as.Date(data$SeriesInd,origin = "1899-12-30")
```

Separate the lines with actual data (1622 x 6 series = 9732) from the lines that are empty, but just give the dates over which we should forecast.

```{r}
dates_to_forecast <- unique(data$SeriesInd[9733:nrow(data)])

data <- data[1:9732,]
```

## Check which dates included in SeriesInd and dates to be forecast.

First, let's get a general overview by seeing how many of each day of the week included.

```{r}
table(weekdays(unique(data$SeriesInd)))
```

Looks like Wednesdays and Thursdays are not included at all. What about other dates missing from SeriesInd?

```{r}
data_range <- as.Date(seq(from=min(data$SeriesInd),to=max(data$SeriesInd),by="days"),origin="1970-01-01")

missing_dates <- as.Date(setdiff(data_range,unique(data$SeriesInd)),origin="1970-01-01")

missing_dates <- missing_dates[weekdays(missing_dates) %in% c("Monday","Tuesday","Friday","Saturday","Sunday")]

missing_dates <- missing_dates[order(lubridate::month(missing_dates),day(missing_dates))]

data.frame(Day = weekdays(missing_dates),
	Date = missing_dates)
```

We get the following list of holidays.

* First Friday of January
* One time holiday in 2014 on Friday 02/28/14 and Saturday 03/01/14.
* Last Monday in March
* Late April, exact date varies (has been April 24 to 26 so far)
* Early May, exact date varies (has been May 2 or 3 so far)
* One Friday a month (toward the middle or end of the month) every year in May, June, and September
* One Tuesday a year from late July to late August
* Early November, exact date varies (has been November 1 to 4 so far)

Now, what range of dates are we being asked to forecast? And does this include any holidays?

```{r, echo=FALSE}
print("Date range to forecast, and weekdays of forecast dates:")
```

```{r}
range(dates_to_forecast)
table(weekdays(dates_to_forecast))
```

```{r, echo=FALSE}
print("Dates missing from to be forecast:")
```

```{r}
forecast_range <- as.Date(seq(from=min(dates_to_forecast),to=max(dates_to_forecast),by="days"),origin="1970-01-01")

missing_dates <- as.Date(setdiff(forecast_range,dates_to_forecast),origin="1970-01-01")

missing_dates <- missing_dates[weekdays(missing_dates) %in% c("Monday","Tuesday","Friday","Saturday","Sunday")]

missing_dates <- missing_dates[order(lubridate::month(missing_dates),day(missing_dates))]

data.frame(Day = weekdays(missing_dates),
        Date = missing_dates)
```

Looks like we are not being asked to forecast for Wednesdays or Thursdays, nor on any holidays. So that's good.

## Convert to long format and select relevant series.

Let's convert from wide to long format, and select only our series of interest.

```{r}
data_gathered <- gather(data,
        key="Variable",
        value="Value",
        -SeriesInd,-group)
data_gathered <- data.frame(data_gathered,
        Group.plus.var = paste0(data_gathered$group,"-",data_gathered$Variable),
        stringsAsFactors=FALSE)
data_gathered <- data_gathered[data_gathered$Group.plus.var %in% c("S02-Var02","S02-Var03","S03-Var05","S03-Var07","S04-Var01","S04-Var02"),]
```

## Check for missing values where date is present in SeriesInd.

Let's print all rows with NA.

```{r}
rows_with_NA <- data_gathered[which(is.na(data_gathered$Value) == TRUE),]
rows_with_NA[order(rows_with_NA$SeriesInd,rows_with_NA$group,rows_with_NA$Variable),]
```

We find that 2017-06-11 and 2017-06-12 are missing for all non-Var02 variables.

2017-09-19 and 2017-09-22 are missing for most non-Var02 variables except for S04-Var01.

## Explore distribution of data in general and over time.

Start by making some simple line plots from the data.

```{r,fig.height=6,fig.width=10}
ggplot(data_gathered,
aes(SeriesInd,Value)) +
geom_line() +
facet_wrap(~Group.plus.var,ncol=3,nrow=2,scales="free") +
xlab("Time")
```

Looks like we may want to try plotting both Var02 variables in log10.

Also, let's see what S03-Var05 and S03-Var07 look like on the same plots.

Finally, plot S02-Var03 with a truncated y-axis.

```{r,fig.height=6,fig.width=8}
data_gathered_Var02 <- data_gathered[data_gathered$Variable == "Var02",]

ggplot(data_gathered_Var02,
aes(SeriesInd,Value)) +
geom_line() +
facet_wrap(~Group.plus.var,ncol=3,nrow=2) +
xlab("Time") +
scale_y_log10()
```

```{r,fig.height=6,fig.width=6}
data_gathered_S03 <- data_gathered[data_gathered$group == "S03",]

ggplot(data_gathered_S03,
aes(SeriesInd,Value,colour=Variable)) +
geom_line() +
xlab("Time") +
theme(legend.position="none") +
ggtitle("S03, Var05 (red) and Var07 (blue)")
```

```{r,fig.height=6,fig.width=6}
ggplot(data_gathered[data_gathered$Group.plus.var == "S02-Var03",],
aes(SeriesInd,Value)) +
geom_line() +
xlab("Time") +
ggtitle("S02-Var03, truncated y-axis") +
coord_cartesian(ylim = c(min(data_gathered[data_gathered$Group.plus.var == "S02-Var03","Value"],na.rm=TRUE),18))
```

From these plots, we find the following notes:

* Both Var02 variables appear much more noisy than the other variables.
* Converting Var02 to log10 seems to be the right choice, making the outliers less extreme.
* S02-Var02 seems like it may have a decreasing trend.
* Both S03 variables appear very similar. If this were an actual business report, I would suggest only using one or combining somehow, as they are mostly redundant.
* S02-Var03 has one strong outlier that should definitely be adjusted.
* All of the non-Var02 variables look like they may have trend, seasonality, and/or a cyclic pattern. Hard to tell from just this graph alone.

Now, let's make a histogram of each variable. Use log10 for Var02.

```{r,fig.height=6,fig.width=10}
variables <- unique(data_gathered$Group.plus.var)[order(unique(data_gathered$Group.plus.var))]

par(mfrow=c(2,3))

for(var in variables)
{
	if(var == "S02-Var02" | var == "S04-Var02"){
		values = log10(data_gathered$Value[data_gathered$Group.plus.var == var])
		title = paste0(var," log10")
		}
	if(var != "S02-Var02" & var != "S04-Var02"){
		values = data_gathered$Value[data_gathered$Group.plus.var == var]
		title = var
		}
	hist(values,
	xlab="Values",
	ylab="Number of dates",
	labels=TRUE,
	main=title)
}
```

Also make a histogram minus the outlier for S02-Var03.

```{r,fig.height=6,fig.width=6}
var = "S02-Var03"

values = data_gathered$Value[data_gathered$Group.plus.var == var]
values = values[values < 20]

title = paste0(var," minus outlier")

hist(values,
xlab="Values",
ylab="Number of dates",
labels=TRUE,
main=title)
```

Notes:

* Both Var02 variables actually look relatively close to normal once you convert to log10.
* Both S03 variables and S02-Var03 minus outlier actually look more like a uniform distribution than normal, other than at the extremes. But at least there is no major skew.
* S04-Var01 has some right skew, although it is not super extreme.

## Explore distribution of differenced data in general and over time.

Let's make the same plots, but taking the difference of each value over the previous value.

For Var02, take the difference of the log10 values.

```{r}
var = variables[1]

data_this_var <- data_gathered[which(is.na(data_gathered$Value) == FALSE & data_gathered$Group.plus.var == var),]

differenced_data_this_var <- data.frame(SeriesInd = data_this_var$SeriesInd[2:nrow(data_this_var)],
	Difference = log10(data_this_var$Value[2:nrow(data_this_var)]) - log10(data_this_var$Value[1:(nrow(data_this_var) - 1)]),
	stringsAsFactors=FALSE)
```

```{r}
differenced_data_this_var <- data.frame(differenced_data_this_var,
	Variable = var,
	stringsAsFactors=FALSE)
```

```{r}
differenced_data <- differenced_data_this_var
```

```{r}
for(var in variables[2:length(variables)])
{
	data_this_var <- data_gathered[which(is.na(data_gathered$Value) == FALSE & data_gathered$Group.plus.var == var),]
	if(var != "S04-Var02")
	{
		differenced_data_this_var <- data.frame(SeriesInd = data_this_var$SeriesInd[2:nrow(data_this_var)],
			Difference = data_this_var$Value[2:nrow(data_this_var)] - data_this_var$Value[1:(nrow(data_this_var) - 1)],
			stringsAsFactors=FALSE)
	}
	if(var == "S04-Var02")
	{
		differenced_data_this_var <- data.frame(SeriesInd = data_this_var$SeriesInd[2:nrow(data_this_var)],
			Difference = log10(data_this_var$Value[2:nrow(data_this_var)]) - log10(data_this_var$Value[1:(nrow(data_this_var) - 1)]),
			stringsAsFactors=FALSE)
	}
	differenced_data_this_var <- data.frame(differenced_data_this_var,
		Variable = var,
		stringsAsFactors=FALSE)
	differenced_data <- rbind(differenced_data,differenced_data_this_var)
}
```

```{r,fig.height=6,fig.width=10}
par(mfrow=c(2,3))

for(var in variables)
{
	if(var == "S02-Var02" | var == "S04-Var02"){
        title = paste0(var," differenced log10 data")
	}
	if(var != "S02-Var02" & var != "S04-Var02"){
		title = paste0(var," differenced data")
	}
	values = differenced_data$Difference[differenced_data$Variable == var]
	hist(values,
		xlab="Values",
		ylab="Number of dates",
        labels=TRUE,
        main=title)
}
```

```{r, fig.height=6,fig.width=8}
par(mfrow=c(2,2))

for(var in c("S02-Var03","S03-Var05","S04-Var01"))
{
	values = differenced_data$Difference[differenced_data$Variable == var] 
	if(var == "S02-Var03"){outlier = 1}
	if(var == "S03-Var05"){outlier = 5}
	if(var == "S04-Var01"){outlier = 1}

	values = values[abs(values) < outlier]
	
	title = paste0(var,"\nDifferenced, minus outliers")

	hist(values,
		xlab="Values",
		ylab="Number of dates", 
		labels=TRUE,
		main=title)
}
```

Let's plot line plots.

```{r,fig.height=6,fig.width=10}
outliers <- which(differenced_data$Variable == "S02-Var03" & abs(differenced_data$Difference) > 10)

ggplot(differenced_data[setdiff(1:nrow(differenced_data),outliers),],
aes(SeriesInd,Difference)) +
geom_line() +
facet_wrap(~Variable,ncol=3,nrow=2,scales="free") +
ylab("Differences (after log10 if Var02)") +
ggtitle("Minus S02-Var03 outlier")
```

Data looks a lot closer to white noise than before.

## Check for day of the week effects.

One "seasonal" pattern that would be hard to see in the overall plot would be day of the week effects.

Let's make sure we do not see any.

```{r,fig.width=10,fig.height=15}
data_gathered_with_day <- data.frame(data_gathered,
	Day = weekdays(data_gathered$SeriesInd),
	stringsAsFactors=FALSE)

data_gathered_with_day$Group.plus.var <- factor(data_gathered_with_day$Group.plus.var,
	levels=variables)

data_gathered_with_day$Day <- factor(data_gathered_with_day$Day,
	levels=c("Friday","Saturday","Sunday","Monday","Tuesday"))

ggplot(data_gathered_with_day,
aes(SeriesInd,Value)) +
geom_line() +
facet_wrap(~Group.plus.var + Day,ncol=5,nrow=6,scales="free")
```

```{r,fig.width=10,fig.height=6}
ggplot(data_gathered_with_day[as.vector(data_gathered_with_day$Group.plus.var) %in% c("S02-Var02","S04-Var02"),],
aes(SeriesInd,Value)) +
geom_line() +
facet_wrap(~Group.plus.var + Day,ncol=5,nrow=2) +
scale_y_log10()
```

Looks like we do not see any day of the week effects.

# Data cleaning and transformation

## Convert Var02 to log10.

Convert both Var02 variables to log10.

```{r}
data_gathered$Value[data_gathered$Group.plus.var %in% c("S02-Var02","S04-Var02")] <- log10(data_gathered$Value[data_gathered$Group.plus.var %in% c("S02-Var02","S04-Var02")])
```

## Outlier removal

### Adjust S02-Var03 outlier to fit in normal distribution.

For the one obvious S02-Var03 outlier, set it equal to the next-highest value.

```{r}
S02_Var03 <- data_gathered$Value[data_gathered$Group.plus.var == "S02-Var03"]
S02_Var03_max <- max(S02_Var03,na.rm=TRUE)
new_value <- max(S02_Var03[S02_Var03 < S02_Var03_max],na.rm=TRUE)

data_gathered$Value[data_gathered$Group.plus.var == "S02-Var03" & data_gathered$Value == S02_Var03_max] <- new_value
```

### Check if need to remove any other outliers.

Let's make a histogram of centered values for each variable now.

```{r,fig.height=6,fig.width=10}
par(mfrow=c(2,3))

for(var in variables)
{
	non_NA <- data_gathered$Value[data_gathered$Group.plus.var == var]
	non_NA <- non_NA[which(is.na(non_NA) == FALSE)]
	non_NA <- (non_NA - mean(non_NA))/sd(non_NA)
	hist(non_NA,
		xlab="Z-score",
		ylab="Number of dates",
		main=var,
		labels=TRUE)
}
```

Let's also adjust the value with z-score > 5 in S04-Var02, then plot before and after.

```{r}
S04_Var02_pre_transform <- data_gathered[data_gathered$Group.plus.var == "S04-Var02",]

S04_Var02_pre_transform$Group.plus.var <- paste0(as.vector(S04_Var02_pre_transform$Group.plus.var),"-before")
```

```{r}
S04_Var02 <- data_gathered$Value[data_gathered$Group.plus.var == "S04-Var02"]
S04_Var02_max <- max(S04_Var02,na.rm=TRUE)
S04_Var02_minus_max <- S04_Var02[which(is.na(S04_Var02) == FALSE & S04_Var02 < S04_Var02_max)]

data_gathered$Value[data_gathered$Group.plus.var == "S04-Var02" & data_gathered$Value == S04_Var02_max] <- max(S04_Var02_minus_max)
```

```{r}
S04_Var02_post_transform <- data_gathered[data_gathered$Group.plus.var == "S04-Var02",]

S04_Var02_post_transform$Group.plus.var <- paste0(as.vector(S04_Var02_post_transform$Group.plus.var),"-post")
```

```{r}
S04_Var02 <- rbind(S04_Var02_pre_transform,S04_Var02_post_transform)
```

```{r}
ggplot(S04_Var02,
aes(SeriesInd,Value)) +
geom_line() +
facet_wrap(~Group.plus.var) +
ggtitle("Before and after outlier adjust")
```

Looks good!

### Filling in missing dates

I think Nathan's strategy to use na.kalman with auto.arima model was a good one here. Let's do that.

```{r}
data_range <- as.Date(seq(from=min(data$SeriesInd),
	to=max(data$SeriesInd),
	by="days"),
	origin="1970-01-01")

missing_dates <- as.Date(setdiff(data_range,unique(data$SeriesInd)),
	origin="1970-01-01")

data_gathered <- rbind(data_gathered,
	data.frame(SeriesInd = rep(missing_dates,times=6),
		group = rep(unlist(lapply(strsplit(variables,"-"),"[[",1)),each=length(missing_dates)),
		Variable = rep(unlist(lapply(strsplit(variables,"-"),"[[",2)),each=length(missing_dates)),
		Value = NA,
		Group.plus.var = rep(variables,each=length(missing_dates)),
		stringsAsFactors=FALSE))

data_gathered <- data_gathered[order(data_gathered$Group.plus.var,data_gathered$SeriesInd),]
```

```{r}
data_gathered_as_list <- vector("list",length=length(variables))

for(i in 1:length(variables))
{
var <- variables[i]
data_gathered_as_list[[i]] <- data_gathered$Value[data_gathered$Group.plus.var == var]
}

names(data_gathered_as_list) <- variables
```

```{r}
data_gathered_as_list <- lapply(data_gathered_as_list,FUN=function(x)na.kalman(x,model="auto.arima"))
```

```{r,eval=TRUE}
data_gathered <- data.frame(data_gathered[,c("SeriesInd","group","Variable")],
	Value = unlist(data_gathered_as_list),
	Group.plus.var = data_gathered$Group.plus.var,
	stringsAsFactors=FALSE)
```

Now, just re-plot to check that everything looks OK.

```{r,fig.height=6,fig.width=10,eval=TRUE}
ggplot(data_gathered,
aes(SeriesInd,Value)) +
geom_line() +
facet_wrap(~Group.plus.var,ncol=3,nrow=2,scales="free") +
xlab("Time")
```

And double check dates match up to the whole data range now.

```{r,eval=TRUE}
length(data_range)

table(weekdays(data_range))

length(unique(data_gathered$SeriesInd))

table(weekdays(unique(data_gathered$SeriesInd)))
```

Looks good!

We can now create a time series for each variable from this point on.

Another thing we may want to do is set aside training vs. test data.

Use October 14, 2016 and onward as the test dates, and everything before as the training data.

```{r}
test_dates <- unique(data_gathered$SeriesInd)[unique(data_gathered$SeriesInd) >= as.Date("2016-10-14",origin="1970-01-01")]

training_data <- data_gathered[data_gathered$SeriesInd < as.Date("2016-10-14",origin="1970-01-01"),]
```

# Forecasting

First, start by running a simple ARIMA model for each variable.

First, run on all dates, not just what we set aside as training data.

Then, can also try training/test split.

Start with S02-Var02.

```{r}
model_variable <- function(myvalues){
	myts <- ts(myvalues,
		start = c(2011,126),
		frequency=365.25)

	autoplot(myts)

	mymodel <- auto.arima(myts)

	return(mymodel)
}
```

```{r}
first_model <- model_variable(data_gathered$Value[data_gathered$Group.plus.var == "S02-Var02"])

first_model
```
