---
title: pH Prediction Exploration and Transformation
author: Heather Geiger
output:
 html_document:
  smart: false
---

# Libraries

```{r, message=FALSE,warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(gridExtra)
library(VennDiagram)
library(corrplot)
```

# Exploratory analysis

## Read in data.

Read in CSV files, which are just the CSV versions of the original Excel sheets.

```{r}
training <- read.csv("pH_prediction_training_data.csv",header=TRUE,stringsAsFactors=FALSE)
test <- read.csv("pH_prediction_test_data.csv",header=TRUE,stringsAsFactors=FALSE)
```

Set aside target (PH) from training into a separate variable. Remove entirely from test.

Then, combine training and test into one data frame.

```{r}
training_target <- training$PH

training <- training[,setdiff(colnames(training),"PH")]
test <- test[,colnames(training)]

alldata <- data.frame(rbind(training,test),
    Data = rep(c("Training","Test"),times=c(nrow(training),nrow(test))),
    stringsAsFactors=FALSE)
```

## Within variables

### Unique values per variable

One useful quick thing to check can be the number and type of unique values per variable.

If for example there are only 12 unique values, and those values are 1-12, then you know that the variable may represent counts rather than a continuous numeric range.

```{r}
unique_per_var <- apply(alldata,2,function(x)length(unique(x)))
```

```{r,echo=FALSE}
print("List of number of unique values per variable:")
```

```{r}
unique_per_var[order(unique_per_var)]
```

```{r,echo=FALSE}
print("Count of unique values per variable for select variables:")
```

```{r}
vars_with_relatively_few_unique <- colnames(alldata)[order(unique_per_var)[2:6]]

apply(alldata[,vars_with_relatively_few_unique],2,function(x)table(x,useNA="ifany"))
```

Looks like brand is blank for a significant number of observations.

In Pressure.Setpoint, the vast majority of values are multiples of 2. So the ones that are between multiples (45.2, 46.4, 46.6, and 46.8) are unusual.

Similarly in Bowl.Setpoint, 122, 126, and 134 are unusual in being not multiples of 10.

### Zeros per variable

Sometimes variables will have a distribution where there are many zeros, but then the nonzero part of the distribution looks like a relatively standard continuous numeric variable.

Let's see if this is the case for any of the variables here.

```{r}
num_zeros_per_var <- rep(0,times=ncol(alldata))

for(i in 1:ncol(alldata))
{
    num_zeros_per_var[i] <- length(which(is.na(alldata[,i]) == FALSE & alldata[,i] == 0))
}

names(num_zeros_per_var) <- colnames(alldata)

num_zeros_per_var[num_zeros_per_var > 0]
```

We find the Hyd.Pressure variables have a very large number of observations equal to 0.

### Repeated values per variable

Any notable often-repeated nonzero values?

```{r}
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
```

```{r}
nonzero_mode_per_var <- data.frame(Variable = colnames(alldata),
    Num.unique = unique_per_var,
    Most.common.nonzero.value = rep(NA,times=ncol(alldata)),
    Num.obs = rep(NA,times=ncol(alldata)),
    stringsAsFactors=FALSE,check.names=FALSE,
    row.names=colnames(alldata))

for(i in 1:ncol(alldata))
{
    mymode <- Mode(alldata[alldata[,i] != 0,i])
    nonzero_mode_per_var$Most.common.nonzero.value[i] <- mymode
    nonzero_mode_per_var$Num.obs[i] <- length(which(alldata[,i] == mymode))
}
```

```{r}
nonzero_mode_per_var <- nonzero_mode_per_var[nonzero_mode_per_var$Num.unique >= 30,]
nonzero_mode_per_var <- nonzero_mode_per_var[which(is.na(nonzero_mode_per_var$Most.common.nonzero.value) == FALSE),]

nonzero_mode_per_var[order(nonzero_mode_per_var$Num.obs,decreasing=TRUE),2:4]
```

Get table for top most common values for a few of these variables.

```{r}
for(var in c("Mnf.Flow","Hyd.Pressure2","Hyd.Pressure3"))
{
    print(var)
    freq_per_var <- data.frame(table(alldata[,var]))
    print(freq_per_var[order(freq_per_var$Freq,decreasing=TRUE)[1:5],])
}
```

We find that Mnf.Flow most common values are -100.2/-100 (together over half of observations), then 0.2 (92 observations).

In addition to the zeros, Hyd.Pressure 2 and 3 each have another repeated value (0.2 for Hyd.Pressure2, -1.2 for Hyd.Pressure3).

### Missing values

How many missing values do we find per variable?

```{r}
missing_per_var <- rep(0,times=ncol(alldata))

for(i in 1:ncol(alldata))
{
    missing_per_var[i] <- length(which(is.na(alldata[,i]) == TRUE))
}

names(missing_per_var) <- colnames(alldata)

missing_per_var[order(missing_per_var)]
```

We find most variables are missing in at least a few observations. Including, brand code should maybe have some as well, once we convert the blanks to NA.

How many missing variables do we tend to find in a given observation?

```{r}
missing_per_obs <- apply(alldata,1,function(x)length(which(is.na(x) == TRUE)))

table(missing_per_obs)
```

Most observations are only missing data for at most one or two variables, which is good.

### Overall distribution of values

Let's make a simple histogram (or barplot if categorical) per variable.

```{r}
alldata$Brand.Code[alldata$Brand.Code == ""] <- "Empty"
```

```{r, fig.width=12,fig.height=24}
par(mfrow=c(8,4))

barplot(table(alldata$Brand.Code),ylab="Obs")

for(var in setdiff(colnames(alldata),c("Data","Brand.Code")))
{
    hist(alldata[,var],ylab="Obs",xlab="",main=var,labels=TRUE)
}
```

For a few of the variables that had zeros, let's also plot original histogram side-by-side with histogram of nonzero values.

Also remove the one very low outlier in Hyd.Pressure.

```{r, fig.width=12,fig.height=6}
par(mfrow=c(2,4))

var = "Balling.Lvl"

hist(alldata[,var],ylab="Obs",xlab="",main=var,labels=TRUE)
hist(alldata[alldata[,var] != 0,var],ylab="Obs",xlab="",main=paste0(var," != 0"),labels=TRUE)

for(var in c("Hyd.Pressure1","Hyd.Pressure2","Hyd.Pressure3"))
{
    hist(alldata[alldata[,var] > -40,var],ylab="Obs",xlab="",main=paste0(var," > -40"),labels=TRUE)
    hist(alldata[alldata[,var] != 0 & alldata[,var] > -40,var],ylab="Obs",xlab="",main=paste0(var," != 0, > -40"),labels=TRUE)
}
```

We see variables with various patterns including bimodal or multimodal, varying degrees of skew, and a few particular values being especially common.

Looks like for Hyd.Pressure 1,2, and 3, most values are much larger than zero if they are not exactly equal to 0.

## Between variables

### Co-occurence of zeros?

One quick obvious thing to check is the correlation between the different Hyd.Pressure variables.

First, do the zeros often co-occur?

```{r}
zero_hyd1 <- which(alldata[,"Hyd.Pressure1"] == 0 & is.na(alldata[,"Hyd.Pressure1"]) == FALSE)
zero_hyd2 <- which(alldata[,"Hyd.Pressure2"] == 0 & is.na(alldata[,"Hyd.Pressure2"]) == FALSE)
zero_hyd3 <- which(alldata[,"Hyd.Pressure3"] == 0 & is.na(alldata[,"Hyd.Pressure3"]) == FALSE)

list_for_venn <- list(Hyd.1 = zero_hyd1,Hyd.2 = zero_hyd2,Hyd.3 = zero_hyd3)

object_for_venn <- venn.diagram(list_for_venn,main="Observations with var = 0",filename=NULL)

grid.draw(object_for_venn)
```

Yes, it seems like they definitely do!

One more thing - let's just look a bit more in detail at the observations where Hyd.Pressure2 = 0.2 and/or Hyd.Pressure3 = -1.2.

Have a feeling at least some of these should be 0.

```{r}
length(which(alldata$Hyd.Pressure2 == 0.2))
length(which(alldata$Hyd.Pressure2 == 0.2 & alldata$Hyd.Pressure1 == 0 & alldata$Hyd.Pressure3 == 0))
```

```{r}
length(which(alldata$Hyd.Pressure3 == -1.2))
length(which(alldata$Hyd.Pressure3 == -1.2 & alldata$Hyd.Pressure1 == 0))
```

Also, let's print the observations where:

* Hyd.Pressure 1 and 2 are 0, but 3 is not (1 observation)
* Hyd.Pressure 3 is 0, but 1 is not (2 observations)
* Hyd.Pressure 1 is 0, but 3 is not, and 3 is also not -1.2 (32 - 26 = 6 observations)

```{r}
myindices <- which(alldata$Hyd.Pressure1 == 0 & alldata$Hyd.Pressure2 == 0 & alldata$Hyd.Pressure3 != 0)

alldata[myindices,paste0("Hyd.Pressure",1:3)]

myindices <- which(alldata$Hyd.Pressure3 == 0 & alldata$Hyd.Pressure1 != 0)

alldata[myindices,paste0("Hyd.Pressure",1:3)]

zero_hyd1 <- which(alldata$Hyd.Pressure1 == 0)
zero_or_neg1.2_hyd3 <- which(alldata$Hyd.Pressure3 == 0 | alldata$Hyd.Pressure3 == -1.2)

myindices <- setdiff(zero_hyd1,zero_or_neg1.2_hyd3)

alldata[myindices,paste0("Hyd.Pressure",1:3)]
```

Looks like we may want to convert 0.2 to 0 for Hyd.Pressure2 when both Hyd.Pressure 1 and 3 are 0.

Same idea for Hyd.Pressure3. Convert -1.2 to 0 when Hyd.Pressure 1 is 0.

Finally, convert 0.2 to 0 for Hyd.Pressure 1 and 2 when 3 = 0.

Oh, and convert NA to 0 when 2/3 are NA and 1 is 0.

Not 100% sure these conversions are correct, so let's also set aside the original data.

```{r}
alldata_original <- alldata

alldata[which(alldata$Hyd.Pressure2 == 0.2 & alldata$Hyd.Pressure1 == 0 & alldata$Hyd.Pressure3 == 0),"Hyd.Pressure2"] <- 0
alldata[which(alldata$Hyd.Pressure3 == -1.2 & alldata$Hyd.Pressure1 == 0),"Hyd.Pressure3"] <- 0

myindices <- which(alldata$Hyd.Pressure1 == 0.2 & alldata$Hyd.Pressure2 == 0.2 & alldata$Hyd.Pressure3 == 0)

alldata[myindices,"Hyd.Pressure1"] <- 0
alldata[myindices,"Hyd.Pressure2"] <- 0

myindices <- which(alldata$Hyd.Pressure1 == 0 & is.na(alldata$Hyd.Pressure2) == TRUE & is.na(alldata$Hyd.Pressure3) == TRUE)

alldata[myindices,"Hyd.Pressure2"] <- 0
alldata[myindices,"Hyd.Pressure3"] <- 0

myindices <- which(alldata$Hyd.Pressure1 == 0 & alldata$Hyd.Pressure3 == 0 & alldata$Hyd.Pressure2 == 0.2)

alldata[myindices,"Hyd.Pressure2"] <- 0
```

```{r,echo=FALSE,eval=FALSE}
#Making Venn to double check filtering went correctly.

zero_hyd1 <- which(alldata[,"Hyd.Pressure1"] == 0 & is.na(alldata[,"Hyd.Pressure1"]) == FALSE)
zero_hyd2 <- which(alldata[,"Hyd.Pressure2"] == 0 & is.na(alldata[,"Hyd.Pressure2"]) == FALSE)
zero_hyd3 <- which(alldata[,"Hyd.Pressure3"] == 0 & is.na(alldata[,"Hyd.Pressure3"]) == FALSE)

list_for_venn <- list(Hyd.1 = zero_hyd1,Hyd.2 = zero_hyd2,Hyd.3 = zero_hyd3)

object_for_venn <- venn.diagram(list_for_venn,main="Observations with var = 0 after clean",filename=NULL)

grid.draw(object_for_venn)
```

```{r,echo=FALSE,eval=FALSE}
#Venn did not show the following:
#One obs with 0 in 1, but not 2 or 3
#One obs with 0 in 1 and 2, but not 3

#Make sure these are still there.

myindices <- which(alldata$Hyd.Pressure1 == 0 & alldata$Hyd.Pressure2 != 0 & alldata$Hyd.Pressure3 != 0)

alldata[myindices,paste0("Hyd.Pressure",1:3)]

myindices <- which(alldata$Hyd.Pressure1 == 0 & alldata$Hyd.Pressure2 == 0 & alldata$Hyd.Pressure3 != 0)

alldata[myindices,paste0("Hyd.Pressure",1:3)]
```

Now, just curious, what do the other two variables look like when Hyd.Pressure2 is 0.2 or Hyd.Pressure3 is -1.2?

```{r}
myindices1 <- which(alldata$Hyd.Pressure2 == 0.2)

myindices2 <- which(alldata$Hyd.Pressure3 == -1.2)

length(myindices1)
length(myindices2)

length(intersect(myindices1,myindices2))

hyd1_when_hyd2plus3_repeated_values <- as.numeric(as.vector(alldata[myindices1,"Hyd.Pressure1"]))

table(hyd1_when_hyd2plus3_repeated_values[order(hyd1_when_hyd2plus3_repeated_values)])
```

After some transformations in other rows to 0, looks like Hyd.Pressure2 = 0.2 and Hyd.Pressure3 = -1.2 always co-occur.

Then Hyd.Pressure1 is within a range of different low values.

### Correlations between certain specific variables

Next, make scatterplots looking at correlation including nonzero values for Hyd.Pressure 1-4.

```{r,fig.width=12,fig.height=12}
pairs(alldata[alldata[,"Hyd.Pressure1"] > -40,paste0("Hyd.Pressure",1:4)],lower.panel=NULL,main="Minus 1/2/3 < -40 observation")
```

Other than the common zeros, the Hyd.Pressure variables are not quite as correlated as I expected.

Some correlation between 2 and 3, but it is not that strong other than when both are very low.

### Correlations between all predictors

Now, time to take an unbiased look and see which variables are most correlated.

For now, exclude Hyd.Pressure1-3 and Mnf.Flow from this since they have very oft-repeated values. Also exclude the categorical variables.

```{r,fig.width=6,fig.height=6}
alldata_for_correlations <- alldata[,setdiff(colnames(alldata),c("Data","Brand.Code","Hyd.Pressure1","Hyd.Pressure2","Hyd.Pressure3","Mnf.Flow"))]

corrplot(cor(alldata_for_correlations,use="pairwise.complete.obs"),type="upper",diag=FALSE)
```

We find the following very strong correlations:

* Carb.Volume with Density, Balling, Alch.Rel, Carb.Rel, and Balling.Lvl (and correlations within these as well)
* Carb.Pressure with Carb.Temp
* Filler.Level with Bowl.Setpoint
* Filler.Speed with MFR

Let's make some scatterplots!

```{r,fig.width=12,fig.height=12}
pairs(alldata[,c("Carb.Volume","Density","Balling","Alch.Rel","Carb.Rel","Balling.Lvl")],lower.panel=NULL)
```

```{r}
par(mfrow=c(2,2))

plot(alldata$Balling,alldata$Density,xlab="Balling",ylab="Density",
    ylim=c(-.20,max(alldata$Density,na.rm=TRUE)))

model1_indices <- which(alldata$Density > (0.4*alldata$Balling))
model2_indices <- which(alldata$Density <= (0.4*alldata$Balling))
model2_indices <- setdiff(model2_indices,which(alldata$Density == min(alldata$Density,na.rm=TRUE)))

abline(lm(Density ~ Balling,data=alldata[model1_indices,]),lty=2,col="red")
abline(lm(Density ~ Balling,data=alldata[model2_indices,]),lty=2,col="blue")

legend("topleft",
    legend=c("y = .40x+.32","y = .40x-.17"),
    col=c("red","blue"),lty=2,bty="n",cex=0.75)

plot(alldata$Carb.Pressure,alldata$Carb.Temp,xlab="Carb.Pressure",ylab="Carb.Temp")
plot(alldata$Filler.Level,alldata$Bowl.Setpoint,xlab="Filler.Level",ylab="Bowl.Setpoint")
plot(alldata$Filler.Speed,alldata$MFR,xlab="Filler.Speed",ylab="MFR")
```

Looks like density and balling are actually almost perfectly correlated. Except, some points are modelled by a different relationship than others.

For the remaining variables, the correlation coefficient generally makes sense with the scatterplot, but not always as expected. Sometimes, there is just a correlation when points from one modal peak of variable1 have higher values for variable2 than points from the other (lower) modal peak of variable1.

Now, let's look at brand code vs. other variables.

```{r,fig.width=12,fig.height=6}
brand_code <- alldata$Brand.Code

brand_code_dat <- data.frame(A = ifelse(alldata$Brand.Code == "A",1,0),
    B = ifelse(alldata$Brand.Code == "B",1,0),
    C = ifelse(alldata$Brand.Code == "C",1,0),
    D = ifelse(alldata$Brand.Code == "D",1,0),
    stringsAsFactors=FALSE)

correlations_with_brand <- cor(brand_code_dat,alldata_for_correlations,use="pairwise.complete.obs")

correlations_with_brand_max <- apply(correlations_with_brand,2,function(x)max(abs(x)))

vars_corr_with_brand <- colnames(correlations_with_brand)[correlations_with_brand_max > 0.4]

par(mfrow=c(2,4))

for(var in vars_corr_with_brand)
{
    boxplot(alldata[,var] ~ alldata$Brand.Code,ylab=var)
}
```

We find the following patterns.

* Higher Carb.Volume, Carb.Pressure (effect not as strong), Density, Balling, Alch.Rel, Carb.Rel, and Balling.Lvl in brands A and D.
* Lower Hyd.Pressure4 in brand D.

For Hyd.Pressure 1-3, let's see if any patterns for where Hyd.Pressure1 is 0 vs. not.

```{r,fig.width=12,fig.height=6}
correlations_with_hyd_zero <- cor(ifelse(alldata$Hyd.Pressure1 == 0,1,0),alldata_for_correlations,use="pairwise.complete.obs")

vars_corr_with_hyd_zero <- colnames(correlations_with_brand)[abs(correlations_with_hyd_zero) >= 0.4]

zero_indices <- which(alldata$Hyd.Pressure1 == 0)
nonzero_indices <- setdiff(1:nrow(alldata),zero_indices)

par(mfrow=c(2,4))

for(var in vars_corr_with_hyd_zero)
{
    if(var == "Filler.Level"){mybreaks=seq(from=50,to=170,by=10)}
    if(var == "Pressure.Vacuum"){mybreaks=seq(from=-6.75,to=-3.5,by=0.25)}
    if(var == "Bowl.Setpoint"){mybreaks=seq(from=70,to=140,by=10)}
    if(var == "Pressure.Setpoint"){mybreaks=seq(from=42,to=54,by=2)}
    hist(alldata[nonzero_indices,var],xlab=var,ylab="Obs",main="Hyd.Pressure1 != 0",breaks=mybreaks)
    hist(alldata[zero_indices,var],xlab=var,ylab="Obs",main="Hyd.Pressure1 == 0",breaks=mybreaks)
    if(var == "Pressure.Setpoint")
    {
        print(var)
        print(table(alldata[nonzero_indices,var]))
        print(table(alldata[zero_indices,var]))
    }
}
```

We find Filler.Level and Pressure.Vacuum are both higher when Hyd.Pressure1 = 0.

For Bowl.Setpoint and Pressure.Setpoint, we find that the values tend to be less on the extremes and more toward more common values when Hyd.Pressure1 = 0.

Another question - are Hyd.Pressure 1/2/3 correlated with other variables when nonzero?

```{r}
alldata_for_correlations <- alldata[nonzero_indices,setdiff(colnames(alldata),c("Data","Brand.Code","Mnf.Flow"))]

correlations_with_hyd_nonzero <- cor(alldata_for_correlations,use="pairwise.complete.obs")

correlations_with_hyd_nonzero <- correlations_with_hyd_nonzero[paste0("Hyd.Pressure",1:3),setdiff(colnames(correlations_with_hyd_nonzero),paste0("Hyd.Pressure",1:3))]

max_correlations_with_hyd_nonzero <- apply(correlations_with_hyd_nonzero,2,function(x)max(abs(x)))

correlations_with_hyd_nonzero[,max_correlations_with_hyd_nonzero >= 0.4]
```

```{r,fig.width=12,fig.height=8}
par(mfrow=c(2,3))

plot(alldata$Hyd.Pressure1,alldata$Carb.Pressure1,xlab="Hyd.Pressure1",ylab="Carb.Pressure1")
plot(alldata$Hyd.Pressure1,alldata$Filler.Level,xlab="Hyd.Pressure1",ylab="Filler.Level")
plot(alldata$Hyd.Pressure2,alldata$Filler.Speed,xlab="Hyd.Pressure2",ylab="Filler.Speed")
plot(alldata$Hyd.Pressure2,alldata$MFR,xlab="Hyd.Pressure2",ylab="MFR")
plot(alldata$Hyd.Pressure1,alldata$Bowl.Setpoint + rnorm(nrow(alldata),mean=0,sd=0.5),xlab="Hyd.Pressure1",ylab="Bowl.Setpoint")
```

Finally, look at Mnf.Flow at the repeated values vs. others.

```{r}
mnf_flow <- alldata$Mnf.Flow

mnf_flow <- plyr::mapvalues(mnf_flow,
    from=c(-100.2,-100,0.2,setdiff(unique(mnf_flow),c(-100.2,-100,0.2))),
    to=c("-100.2","-100","0.2",rep("Other",times=length(unique(mnf_flow)) - 3)))

mnf_flow_dat <- data.frame(Neg.100point2 = ifelse(mnf_flow == "-100.2",1,0),
    Neg.100 = ifelse(mnf_flow == "-100",1,0),
    Point2 = ifelse(mnf_flow == "0.2",1,0),
    stringsAsFactors=FALSE)

alldata_for_correlations <- alldata[,setdiff(colnames(alldata),c("Data","Brand.Code","Mnf.Flow"))]

correlations_with_mnf_flow <- cor(mnf_flow_dat,alldata_for_correlations,use="pairwise.complete.obs")

correlations_with_mnf_flow_max <- apply(correlations_with_mnf_flow,2,function(x)max(abs(x)))

vars_corr_with_mnf_flow <- colnames(correlations_with_mnf_flow)[correlations_with_mnf_flow_max >= 0.4]

par(mfrow=c(2,2))

for(var in vars_corr_with_mnf_flow)
{
    boxplot(alldata[,var] ~ factor(mnf_flow),ylab=var,xlab="Mnf.Flow")
}

par(mfrow=c(1,1))

boxplot(alldata[,"Hyd.Pressure1"] ~ factor(mnf_flow),ylab="Hyd.Pressure1",xlab="Mnf.Flow")
```

Looks like the zeros in Hyd.Pressure 1/2/3 frequently match up to Mnf.Flow <= 0.2.

Mnf.Flow 0.2 is associated with much lower Filler.Speed.

The lower repeated Mnf.Flow values are associated with higher Pressure.Vacuum.

```{r}
non_repeated_indices <- which(rowSums(mnf_flow_dat) == 0)

alldata_for_correlations <- alldata[non_repeated_indices,setdiff(colnames(alldata),c("Data","Brand.Code","Mnf.Flow"))]

correlations_with_mnf_flow <- cor(alldata$Mnf.Flow[non_repeated_indices],alldata_for_correlations,use="pairwise.complete.obs")

vars_corr_with_mnf_flow <- colnames(alldata_for_correlations)[abs(correlations_with_mnf_flow) >= 0.4]

vars_corr_with_mnf_flow

par(mfrow=c(2,3))

for(var in vars_corr_with_mnf_flow)
{
    if(var != "MFR" & var != "Oxygen.Filler")
    {
        plot(alldata[,var],alldata$Mnf.Flow,xlab=var,ylab="Mnf.Flow",ylim=c(min(alldata$Mnf.Flow[alldata$Mnf.Flow > 0.2],na.rm=TRUE),max(alldata$Mnf.Flow,na.rm=TRUE)))
    }
    if(var == "MFR")
    {
        plot(alldata[,var],alldata$Mnf.Flow,xlab=var,ylab="Mnf.Flow",ylim=c(min(alldata$Mnf.Flow[alldata$Mnf.Flow > 0.2],na.rm=TRUE),max(alldata$Mnf.Flow,na.rm=TRUE)),
        xlim=c(600,850))
    }
    if(var == "Oxygen.Filler")
    {
        plot(alldata[,var],alldata$Mnf.Flow,xlab=var,ylab="Mnf.Flow",ylim=c(min(alldata$Mnf.Flow[alldata$Mnf.Flow > 0.2],na.rm=TRUE),max(alldata$Mnf.Flow,na.rm=TRUE)),
        xlim=c(0,0.1))
    }
}
```

Looks like we see some of the same variables that were correlated with Mnf.Flow as a factor, along with a few new ones.

## Variables vs. target

Separate out training data.

```{r}
training <- alldata[alldata$Data == "Training",]
```

Start with a simple boxplot for brand code.

```{r}
boxplot(training_target ~ training$Brand.Code,ylab="pH")
```

Looks like brand C may have a somewhat lower pH.

Plot scatterplots for remaining variables.

```{r,fig.width=12,fig.height=24}
par(mfrow=c(8,4))

for(var in setdiff(colnames(training),c("Data","Brand.Code")))
{
    plot(training[,var],training_target,xlab=var,ylab="pH",main=paste0(var," cor=",round(cor(training[,var],training_target,use="pairwise.complete.obs"),digits=2)))
}
```

Looks like there are a fair number of variables that will probably not end up in the final model, as they are not really associated with the target (pH) at all.

There also seem like there may be some non-linear associations. For example, Usage.cont generally has constant pH throughout its range, except some observations with Usage.cont > 23 or so have much lower pH than the other points.
